Stability Analysis of Recurrent Neural Networks \\ using Ordinary Differential Equations

### 2020-09-23 p5/stability_rnn.tex
WIP 23Sep2020

\documentclass{article}
\usepackage{amsmath, amssymb, amsfonts, amsthm}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\smax}{softmax}
\numberwithin{equation}{section}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem*{rem}{Remark}
\begin{document}
\title{Stability Analysis of Recurrent Neural Networks \\
    using Ordinary Differential Equations}
\author{Author's name}
\maketitle
\section{Introduction}\label{s1}
Recurrent neural networks feed their output to their input. This feature
makes them look like algorithms to get numerical solutions of ordinary
differential equations. It is well known that training neural networks is
rife with vanishing and exploding gradients\cite{gereon2018hands}. This
phenomenon is closely related to the behaviour of the numerical algorithms
to solve ordinary differential equations, a topic that has been studied
intensively for several decades. We use this analogy to bank upon the 
vast research about ordinary differential equations to derive criteria 
for the stability of recurrent neural networks.

@@

A gated recurrent unit with the structure
\begin{eqnarray}
z_t &=& \sigma(W_zx_t + U_zh_{t-1} + b_z) \label{s11e3} \\
r_t &=& \sigma(W_rx_t + U_rh_{t-1} + b_r) \label{s11e4} \\
h_t &=& (1 - z_t)\circ\tanh(W_hx_t + U_h(r_t\circ h_{t-1} + b_h)) + 
        z_t \circ h_{t-1} \label{s11e5}
\end{eqnarray}
where $W_z, W_r, W_h \in \mathbb{R}^{d \times p}$, $U_z, U_r, U_h \in 
\mathbb{R}^{d \times d}$ are parameter matrices and  $b_z, b_r, b_h \in
\mathbb{R}^d$ are bias vectors. In the absence of an input the above
equations become
\begin{eqnarray}
z_t &=& \sigma(U_zh_{t-1} + b_z) \label{s11e6} \\
r_t &=& \sigma(U_rh_{t-1} + b_r) \label{s11e7} \\
h_t &=& (1 - z_t)\circ\tanh(U_h(r_t\circ h_{t-1} + b_h)) + 
        z_t \circ h_{t-1} \label{s11e8}
\end{eqnarray}
One can consider equations \eqref{s11e6} to \eqref{s11e8} to be discrete
forms of the system of ODEs \cite{jordan2019gated}
\begin{eqnarray}
z(t) &=& \sigma(U_zh(t) + b_z) \label{s11e9} \\
r(t) &=& \sigma(U_rh(t) + b_r) \label{s11e10} \\
\dot{h}(t) &=& (1 - z(t))\circ
	\left(-h(t)+\tanh(U_h(r(t)\circ h(t)) + b_h)\right) \label{s11e11},
\end{eqnarray}
where the dot over $h$ denotes its time derivative. We consider a special case
of these equations with the following values of parameters, (first row of 
table I in \cite{jordan2019gated},
\begin{eqnarray}
U_r &=&
\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsfonts, amsthm}
\usepackage{graphicx}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\smax}{softmax}
\numberwithin{equation}{section}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem*{rem}{Remark}
\begin{document}
\title{Stability Analysis of Recurrent Neural Networks \\
    using Ordinary Differential Equations}
\author{Author's name}
\maketitle
\section{Introduction}\label{s1}
Recurrent neural networks feed their output to their input. This feature
makes them look like algorithms to get numerical solutions of ordinary
differential equations. It is well known that training neural networks is
rife with vanishing and exploding gradients\cite{gereon2018hands}. This
phenomenon is closely related to the behaviour of the numerical algorithms
to solve ordinary differential equations, a topic that has been studied
intensively for several decades. We use this analogy to bank upon the 
vast research about ordinary differential equations to derive criteria 
for the stability of recurrent neural networks.

@@

A gated recurrent unit with the structure
\begin{eqnarray}
z_t &=& \sigma(W_zx_t + U_zh_{t-1} + b_z) \label{s11e3} \\
r_t &=& \sigma(W_rx_t + U_rh_{t-1} + b_r) \label{s11e4} \\
h_t &=& (1 - z_t)\circ\tanh(W_hx_t + U_h(r_t\circ h_{t-1} + b_h)) + 
        z_t \circ h_{t-1} \label{s11e5}
\end{eqnarray}
where $W_z, W_r, W_h \in \mathbb{R}^{d \times p}$, $U_z, U_r, U_h \in 
\mathbb{R}^{d \times d}$ are parameter matrices and  $b_z, b_r, b_h \in
\mathbb{R}^d$ are bias vectors. In the absence of an input the above
equations become
\begin{eqnarray}
z_t &=& \sigma(U_zh_{t-1} + b_z) \label{s11e6} \\
r_t &=& \sigma(U_rh_{t-1} + b_r) \label{s11e7} \\
h_t &=& (1 - z_t)\circ\tanh(U_h(r_t\circ h_{t-1} + b_h)) + 
        z_t \circ h_{t-1} \label{s11e8}
\end{eqnarray}
One can consider equations \eqref{s11e6} to \eqref{s11e8} to be discrete
forms of the system of ODEs \cite{jordan2019gated}
\begin{eqnarray}
z(t) &=& \sigma(U_zh(t) + b_z) \label{s11e9} \\
r(t) &=& \sigma(U_rh(t) + b_r) \label{s11e10} \\
\dot{h}(t) &=& (1 - z(t))\circ
	\left(-h(t)+\tanh(U_h(r(t)\circ h(t)) + b_h)\right) \label{s11e11},
\end{eqnarray}
where the dot over $h$ denotes its time derivative. We consider a special case
of these equations in the subsections to follow.

\subsection{Example 1}
We choose the following values of parameters, (first row of table I in 
\cite{jordan2019gated}),
\begin{eqnarray}
U_h &=& 3I_2 \label{s11e12} \\
U_r &=& 0 \label{s11e13} \\
U_z &=& 0 \label{s11e14} \\
b_h &=& 0 \label{s11e15} \\
b_r &=& 0 \label{s11e16} \\
b_z &=& 0 \label{s11e17} 
\end{eqnarray}
For these values $z(t) = r(t) = 1/2(1, 1)^T$ and
\begin{equation}\label{s11e18}
\dot{h}(t) = \frac{1}{2}\left(-h(t) + \tanh\left(\frac{3}{2}h(t)\right)\right)
\end{equation}
The stable fixed points of this equation are shown in figure \ref{f1}.
\begin{figure}[h]
\caption{Stable points}\label{f1}
\includegraphics[scale=0.5]{Ex1_Stable_points.png}
\centering
\end{figure}
The coloured dots indicate the start of the trajectory while the blue square
indicates its end. Trajectories seem to converge on the stable points. The
phase portrait also indicates that the arrows point toward the stable points.

This equation also has an unable point at the origin. The phase portrait has
arrows diverging from the origin. The trajectories follow these directions.
\begin{figure}[h]
\caption{Unstable points}\label{f2}
\includegraphics[scale=0.5]{Ex1_Unstable_point.png}
\centering
\end{figure}
We will now examine the behaviour of this unstable point for a collection of
five such GRUs starting their trajectory in the vicinity of the origin.
In each of these figures the trajectory of the GRUs with attention layer
is shown as a dashed line ending with a red square. The trajectories of the
individual GRUs are shown as solid lines ending at the blue square. The
initial conditions of the GRUs were chosen randomly from a set of $16$ points
on a square centred at the origin and of side $0.1$. Figure \ref{f3} shows
the first set of $6$ samples and figure \ref{f4} shows the remaining $6$
samples.

In each of these samples, the trajectory of the GRUs with attention layer
never goes farther than the individual GRUs. It does veer away from the 
origin but at a slower rate. This observation confirms the theoretical analysis
of the previous sections that an attention layer does not eliminate an 
instability. However, they indicate that the attention layer mitigates it to
a certain extent.

\begin{figure}[h]
\caption{Unstable point with attention layer - set 1}\label{f3}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_0}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_1}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_2}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_3}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_4}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_5}
\end{array}$
\end{center}
\end{figure}

\begin{figure}[h]
\caption{Unstable point with attention layer - set 2}\label{f4}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_6}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_7}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_8}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_9}
\end{array}$
\end{center}
\begin{center}$
\begin{array}{cc}
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_10}&
\includegraphics[scale=0.4]{Ex1_Unstable_Point_Sample_9}
\end{array}$
\end{center}
\end{figure}

\subsection{Example 2}
\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}
